---
title: "GB generation share trends"
subtitle: "NG ESO data"
author: "Ben Anderson (b.anderson@soton.ac.uk)"
date: 'Last run at: `r Sys.time()`'
output: 
  bookdown::html_document2:
    fig_caption: yes
    toc: TRUE
    toc_depth: 4
    toc_float: TRUE
    code_folding: hide
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(gridCarbon) # useful functions

myLibs <- c("data.table", # fast data munching
            "lubridate", # time stuff
            "ggplot2", # plots obvs
            "here", # where are we?
            "hms" # more time stuff
            )
gridCarbon::loadLibraries(myLibs) # load the libs (installs if not present)

# local functions ----
makeFlexTable <- function(df, cap = "caption", digits = 1){
  # makes a pretty flextable - see https://cran.r-project.org/web/packages/flextable/index.html
  ft <- flextable::flextable(df)
  ft <- colformat_double(ft, digits = digits)
  ft <- fontsize(ft, size = 9)
  ft <- fontsize(ft, size = 10, part = "header")
  ft <- set_caption(ft, caption = cap)
  return(ft)
}

# parameters ----
dataPath <- path.expand("~/Dropbox/data/UK_NGESO/genMix/") # change to your data loc or for extra points download from https://data.nationalgrideso.com/carbon-intensity1/historic-generation-mix/r/historic_gb_generation_mix#

#dt <- data.table::fread("https://data.nationalgrideso.com/backend/dataset/88313ae5-94e4-4ddc-a790-593554d8c6b9/resource/f93d1835-75bc-43e5-84ad-12472b180a98/download/df_fuel_ckan.csv")
```

# Introduction

Inspired by https://twitter.com/DrSimEvans/status/1508409309775994892

# Data

This analysis uses the [UK NG ESO generation mix](https://data.nationalgrideso.com/carbon-intensity1/historic-generation-mix/r/historic_gb_generation_mix) data. The data also happens to include a half-hourly carbon intensity value.

As far as we can work out this data does _not_ include distributed (i.e. non-grid connected) generation such as small scale wind, solar, hydro, biomass etc which is connected to the LV network. This means the ESO data is likely to _underestimate_ total generation and potentially underestimate the proportion of total generation that is renewable. It is possible that this could be fixed using embedded wind & solar generation data from the [demand update file](https://data.nationalgrideso.com/demand/daily-demand-update).

```{r loadData}
dt_orig <- data.table::fread(paste0(dataPath, "df_fuel_ckan.csv"))

dt_orig[, dv_dateTime := lubridate::as_datetime(DATETIME)] # proper date time

message("Data range from: ", min(dt_orig$dv_dateTime))
message("...to: ", max(dt_orig$dv_dateTime))
```

# Recreating @DrSimEvans' plot

This looks like daily data.

 * Solar + wind (% of gen) vs coal + gas (& of gen))

```{r simEvansData}
dt_orig[, dv_coal_gas_pc := COAL_perc + GAS_perc]
dt_orig[, dv_solar_wind_pc := SOLAR_perc + WIND_perc]
dt_orig[, dv_coal_gas := COAL + GAS]
dt_orig[, dv_solar_wind := SOLAR + WIND]

temp <- dt_orig[, .(dv_dateTime, dv_coal_gas, dv_solar_wind,
                    dv_coal_gas_pc, dv_solar_wind_pc, GENERATION)]

temp[, dv_date := lubridate::date(dv_dateTime)]

plotDT <- temp[,
               .(mean_dv_solar_wind_pc = mean(dv_solar_wind_pc),
                 mean_dv_coal_gas_pc = mean(dv_coal_gas_pc),
                 total_dv_coal_gas = sum(dv_coal_gas),
                 total_dv_solar_wind = sum(dv_solar_wind),
                 total_GENERATION = sum(GENERATION),
                 nObs = .N), # to check for days with < 48 half hours
               keyby = .(dv_date)
               ]
plotDT[, dv_year := lubridate::year(dv_date)]
plotDT[, total_dv_coal_gas_pc := total_dv_coal_gas/total_GENERATION]
plotDT[, total_dv_solar_wind_pc := total_dv_solar_wind/total_GENERATION]

message("Check for days with less than 48 hours - this will be truncated data")
table(plotDT$nObs)
```

Figure \@ref(fig:plotMeanHalfHourPC) shows the mean half-hourly % generation by each type per day. This is slightly convoluted - it is the mean of the sum of the half-hourly XXX_perc values in the original data where XXX is the generation type.

The smoothed curve is estimated for each year via gam. 

```{r plotMeanHalfHourPC, fig.cap="Mean half-hourly % generation by each type per day"}
ggplot2::ggplot(plotDT[dv_year > 2011], aes(x = mean_dv_solar_wind_pc, 
                            y = mean_dv_coal_gas_pc,
                            colour = as.factor(dv_year))) +
  geom_point() +
  geom_smooth() +
  scale_colour_viridis_d(name = "Year") +
  labs(x = "Solar & wind (mean % of half-hourly generation per day)",
       y = "Coal & gas (mean % of half-hourly generation per day)",
       caption = "Data: NG ESO Generation Mix")

ggplot2::ggsave(filename = "GB_plotMeanHalfHourPC.png", 
                path = here::here("rmd","plots"),
                height = 5)
```

Figure \@ref(fig:plotDailyPC) shows the percentage of daily generation by type. This is less convoluted as it is the sum of generation per day for the two categories (solar + wind vs gas + coal) as a % of total daily generation.

Again the smoothed curve is estimated for each year via gam. 

```{r plotDailyPC, fig.cap="Percentage of daily generation by type"}
ggplot2::ggplot(plotDT[dv_year > 2011], aes(x = 100 * total_dv_solar_wind_pc, 
                            y = 100 * total_dv_coal_gas_pc,
                            colour = as.factor(dv_year))) +
  geom_point() +
  geom_smooth() +
  scale_colour_viridis_d(name = "Year") +
  labs(x = "Solar & wind (% of total daily generation)",
       y = "Coal & gas (% of total daily generation)",
       caption = "Data: NG ESO Generation Mix")

ggplot2::ggsave(filename = "GB_plotDailyPC.png", 
                path = here::here("rmd","plots"),
                height = 5)
```
# Half-hourly plots

Just cos we can... helpfully split into 'peak' and 'off peak' periods. Again the smoothed curve is estimated for each year (and demand period) via gam. 

```{r plotHalfHourlyPC, fig.cap="Percentage of half-hourly generation by type"}
dt_orig[, dv_year := lubridate::year(dv_dateTime)]
dt_orig[, dv_hms := hms::as_hms(dv_dateTime)]
# half-hours are the start of the half hours (we think)
dt_orig[, dv_peak := ifelse(dv_hms >= hms::as_hms("07:00:00") & dv_hms < hms::as_hms("09:00:00") , "Morning peak", "Night")]
dt_orig[, dv_peak := ifelse(dv_hms > hms::as_hms("09:00:00") & dv_hms < hms::as_hms("16:00:00") , "Daytime", dv_peak)]
dt_orig[, dv_peak := ifelse(dv_hms >= hms::as_hms("16:00:00") & dv_hms < hms::as_hms("21:00:00") , "Evening peak", dv_peak)]

dt_orig[, dv_hour := lubridate::hour(dv_dateTime)]
# check coding
# table(dt_orig$dv_hour, dt_orig$dv_peak, useNA = "always")

ggplot2::ggplot(dt_orig[dv_year > 2011], aes(x = dv_solar_wind_pc, 
                            y = dv_coal_gas_pc,
                            colour = as.factor(dv_year))) +
  geom_point() +
  facet_wrap(. ~ dv_peak) +
  geom_smooth() +
  scale_colour_viridis_d(name = "Year") +
  labs(x = "Solar & wind (% of half-hourly generation)",
       y = "Coal & gas (% of half-hourly generation)",
       caption = "Data: NG ESO Generation Mix")

ggplot2::ggsave(filename = "GB_plotHalfHourlyPC.png", 
                path = here::here("rmd","plots"),
                height = 5)
```
# Summary

That's it.

# Annex

## Data descriptors

### NGESO gen mix data

```{r skimESO}
skimr::skim(dt_orig)
```

## R environment

Packages etc:

 * base R [@baseR]
 * bookdown [@bookdown]
 * data.table [@data.table]
 * ggplot2 [@ggplot2]
 * here [@here]
 * hms [@hms]
 * knitr [@knitr]
 * lubridate [@lubridate]
 * rmarkdown [@rmarkdown]

# References
