---
params:
  subtitle: ""
  title: ""
  authors: ""
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    self_contained: false
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    fig_width: 5
always_allow_html: yes
bibliography: '`r paste0(here::here(), "/bibliography.bib")`'
---

```{r knitrSetup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # by default turn off code echo to avoid mess
```


```{r rmdSetup, include=FALSE}
# Set start time ----
startTime <- proc.time()

library(gridCarbon)
# Packages used in the report ----
rmdLibs <- c("kableExtra", # fancy tables
             "openair" # very useful plotting functions intended for air quality data analysis
          )
gridCarbon::loadLibraries(rmdLibs)

# Local parameters for .Rmd ----
localParams$genUnits <- "MW"
localParams$ciUnits <- "gCO2e/kWh"
localParams$plotLoc <- paste0(gcParams$repoLoc, "/docs/plots/") # where to put the plots

# Local functions for .Rmd ----

```


# Introduction

Building on [@staffell_measuring_2017] and [@khan_analysis_2018], we are interested in GHG emissions from UK electricity generation over time. We are especially interested in how this might change during the UK covid-19 [lockdown period](https://www.nationalgrideso.com/news/lockdown-effect-tv-viewing-habits-and-electricity-grid) (from `r localParams$lockDownStart`).

Several articles by ythe UK ESO and others have already explored this topic:

 * https://www.nationalgrideso.com/news/lockdown-effect-tv-viewing-habits-and-electricity-grid
 * https://www.nationalgrideso.com/news/planning-easter-during-coronavirus-outbreak
 * https://theconversation.com/we-analysed-electricity-demand-and-found-coronavirus-has-turned-weekdays-into-weekends-134606
 
 
Fortunately the UK Electricity System Operator publishes [half-hourly generation data](https://data.nationalgrideso.com/carbon-intensity1/historic-generation-mix/r/historic_gb_generation_mix) which includes both power generation (and thus 'demand') by fuels and also includes a per-kWh [carbon intensity](https://carbonintensity.org.uk/) for the electricity produced per half-hour. We use this data to explore the following research questions:

 * To what extent has electricty demand shown deviation from 'normal' demand patterns during the lockdown period?

 * Has the composition of fuel sources supplying electriicty changed during this period?

 * Has the lockdown changed greenhouse-gas emissions associated with electriicty generation?

# Data

## Grid generation data {#genData}

This is 'grid' generation from major power stations of various kinds. Data downloaded from ` https://data.nationalgrideso.com/carbon-intensity1/historic-generation-mix/r/historic_gb_generation_mix` and pre-processed.

```{r testGrid}
h <- head(gridGenDT[, .(DATETIME, year, rDateTimeUTC, GENERATION, CARBON_INTENSITY)])

kableExtra::kable(h, caption = "Grid gen data (first 6 rows)") %>%
  kable_styling()
```

Note that according to the dataset [source](https://data.nationalgrideso.com/carbon-intensity1/historic-generation-mix/r/historic_gb_generation_mix):

 * "Data points are either MW or %". This may be `mean MW` over the half hour or it may mean `MWh per half-hour`. It is unclear. The % refers to the fuel mix. Which one could easily calculate from the MW values. But anyway...
 * carbon intensity is helpfully described as _"The carbon intensity of electricity is a measure of how much Carbon dioxide emissions are produced per kilowatt hour of electricity consumed."_. However we assume it is gCO2e/kWh - based on https://carbonintensity.org.uk/ 
 
Table \@ref(tab:descData) shows the mean half hourly generation (MW) and mean carbon intensity over the years covered by the data. It also shows the implied mean half-hourly total kg CO2e per half hour which we have calculated as follows:
{#kgcalc}
 * convert `GENERATION` (`r localParams$genUnits`) to MWh per half hour by `GENERATION/2` (1 MW for half an hour = 1/2 MWh)
 * convert the result to kWh (`* 10000`)
 * multiply by the `CARBON_INTENSITY` which is in `r localParams$ciUnits`
 * divide by 1000 to get Kg
 
Yes, we have *1000 and then /1000 which is 1 but for the sake of clarity we have kept all the steps.

> NB: we are unclear how generation via interconnect is included in the original carbon intensity calculation but note that the [ESO forecast methodology document](https://github.com/carbon-intensity/methodology/raw/master/Carbon%20Intensity%20Forecast%20Methodology.pdf) states that it is.
 
```{r descData}
t <- gridGenDT[, .(sC = sum(C02e_T)/1000000, # convert to Mt for comparison with https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/875508/final-greenhouse-gas-emissions-tables-2018.xlsx
                   mC = mean(C02e_T), # in T
                   mCI = mean(CARBON_INTENSITY),
                   mGen = mean(GW),
                   sGen = sum(GWh)/1000, # TWh
                 maxDate = max(obsDate),
                 nObs = .N), keyby = .(year)]
# gen decrease
ci2017 <- t[year == 2017, sGen]
ci2020 <- t[year == 2020, sGen]
ciDecrease_pc <- round(100*((ci2017 - ci2020)/ci2017))
#ciDecrease_pc
c2017 <- t[year == 2017, sC]
c2020 <- t[year == 2020, sC]
cDecrease_pc <- round(100*((c2017 - c2020)/c2017))

# from https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/875508/final-greenhouse-gas-emissions-tables-2018.xlsx
dbeisDT <- data.table::fread(here::here("publicData", "dbeisDataTo2018.csv"))
dbeisCO2 <- c("DBEIS", "")
mGen <- paste0("Mean halfhourly generation \n (GWh)")
sGen <- paste0("Total annual generation \n (TWh)")
mCI <- paste0("Mean half-hourly carbon intensity \n (" , localParams$ciUnits, ")")
mC <- paste0("Mean halfhourly CO2e \n (tCO2e, calculated)")
sC <- paste0("Total annual CO2e \n(MtCO2e, calculated)")
maxDate <- paste0("Max date in range")

data.table::setnames(t, c("mGen", "sGen", "mCI", "mC","sC", "maxDate"), # old, must be a better way
                     c(mGen,sGen, mCI,mC, sC,maxDate) # new
                     ) # easy way to set human-readable col names
setkey(t, year)
setkey(dbeisDT, year)
t <- dbeisDT[t]
kableExtra::kable(t, digits = 2,
                  caption = "Descriptive stats by year - note that 2020 is incomplete") %>%
  kable_styling()

```

As we can see mean half-hourly generation has declined over the years but much less spectacularly than the mean carbon intensity which decreased by `r ciDecrease_pc` from 2017 to 2020.

Table of peak periods

```{r peakPeriodsTable}
gridGenDT[, month := lubridate::month(rDateTimeUTC)]
t <- gridGenDT[, .(sC = sum(C02e_T)/1000000, # for comparison with https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/875508/final-greenhouse-gas-emissions-tables-2018.xlsx
                   mCI = mean(CARBON_INTENSITY),
                   mGen = mean(GW),
                   sGen = sum(GWh)/1000, # TWh
                   nMonths = uniqueN(month)), keyby = .(year, peakPeriod)]
t[, monthlyTWh := sGen/nMonths]
t[, monthlyC02e_T := sC/nMonths]

wt <- dcast(t, year ~ peakPeriod , value.var = "monthlyTWh")
wt[, Total := `Early morning` + `Morning peak` + `Day time` + `Evening peak` + `Late evening`]
kableExtra::kable(wt, digits = 2,
                  caption = "Monthly total TWh generation by year - note that 2020 is incomplete") %>%
  kable_styling()
```

Fuel trend plots by peak period

```{r fuelTrendsPeakPeriod}

fuelsDT <- gridGenDT[, c("rDateTimeUTC","peakPeriod","COAL","GAS","NUCLEAR","WIND","HYDRO","BIOMASS","SOLAR","STORAGE","IMPORTS","OTHER")]
m <- melt(fuelsDT, id.vars = c("rDateTimeUTC", "peakPeriod"), variable.name = "Fuel", value.name = "MW")
m[, GW := MW/1000]
m[, year := lubridate::year(rDateTimeUTC)]
m[, month := lubridate::month(rDateTimeUTC)]
m[, yearMonth := lubridate::ymd(paste(year,"-", month, "-01"))] # fix

# m[, plotDate := lubridate::ymd(paste(year,"-", month, "-01"))] # fix
m[, q := quarter(rDateTimeUTC)]
m[,plotDate := lubridate::ymd(paste(year,"-", q, "-01"))]
dateLab <- "Quarter"
plotDT <- m[, .(meanGW = mean(GW)), 
            keyby = .(plotDate , Fuel , peakPeriod)]
p <- ggplot2::ggplot(plotDT, aes(x = plotDate, y = meanGW, colour = Fuel)) +
  geom_point() +
  geom_smooth(method = "loess") +
  theme(legend.position = "bottom") + 
  #scale_y_log10() +
  facet_grid(. ~ peakPeriod) +
  labs(x = dateLab,
         y = "Mean half hourly GW")

# yMax <- max(plotDT$meanGW)
# yMin <- min(plotDT$meanGW)
# p <- addLockdownRect(p, 
#                 from = gcParams$UKlockDownStartDate, 
#                 to = gcParams$UKlockDownEndDate, 
#                 yMin = yMin, yMax = yMax)
p
ggplot2::ggsave(here::here("docs", "plots", "ukGWFuelsTrends.png"), 
                width = 10, units = "in",
                p)
```

Repeat using openair::smoothTrend() for de-seasoning and other benefits. NB: deaseasoning produces -ve values for coal :-)

```{r stFuelTrendsPeakPeriod, fog.cap="Mean half-hourly GW"}
fuels <- c("COAL","GAS","NUCLEAR","WIND","HYDRO","BIOMASS","SOLAR","STORAGE","IMPORTS","OTHER")
gridGenDT[, date := obsDate]
oaGWFuels <- openair::smoothTrend(gridGenDT, pollutant = fuels,
                     ylab = "GW", deseason = TRUE,
                  xlab = "Year",
                  date.format = "%Y",
                  date.breaks = 4,
                  type = c("peakPeriod"), 
                  layout = c(5,1),
                  key.columns = 3
                  )

p <- oaGWFuels$plot
png(filename=here::here("docs/plots/openAir_UK_TheilSenoaGWFuelsPlot.png"), 
    width = 960, units = "px")
# dpi = 600, width = 13.333, height = 7.5, units = "in"
plot(p)
dev.off()
```

## Embedded generation data

Essentially 'non-grid' generation from solar photovoltaic and small scale wind which is 'embedded' - i.e. non-grid connected as it is connected 'downstream' of the grid exit points. We are not entirely sure if this is accounted for in the grid dataset or not.

> We have not yet found a source of this data (if we even need it).

> For now embedded generation data is probably _not_ included in the following analysis but we would expect it to depress grid demand when there is greatest insolation (middle of the day, obvs) and wind (largely random in the UK?).

# Analysis

In this section we analyse changes in electricity demand and associated carbon emissions during the UK Covid-19 lockdown via analysis of the generation data. 

## Generation: Analysing deviation from 'normal'

Several [articles](https://theconversation.com/we-analysed-electricity-demand-and-found-coronavirus-has-turned-weekdays-into-weekends-134606) and [analyses](https://twitter.com/DrIAGWilson/status/1255100828781883394) have suggested that demand (and thus generation) patterns have shifted so that [weekdays have become more like weekends](http://www.ukerc.ac.uk/news/covid-lockdown-electricity-demand.html).

### Overall trends

Figure \@ref(fig:TheilSenGWhPlot) uses the openair `TheilSen()` function [@openair] to create a de-seasoned trend plot for mean monthly half-hourly (i.e. the mean of all half-hourly GW values for each month) generation in GW split by peak period (07:00 - 08:30 and 16:00 - 20:00) . 

These plots show that there has been a steady decline of ~ 0.4 - 0.5 GW mean half hourly generation (and thus demand) per year since `r min(gridGenDT$year)` in each time period. The substantial below-trend fall in generation during lockdown in early Spring 2020 is clear, especailly for the morning peak, day-time and, to a lesser extent, evening peak periods. There appears to have been a smaller below-trend reduction at other times as we would expect. We can also see a notable drop in generation in late 2015 which was [unusually warm and wet](https://www.metoffice.gov.uk/binaries/content/assets/metofficegovuk/pdf/weather/learn-about/uk-past-events/state-of-uk-climate/mo-state-of-uk-climate-2015-v3.pdf).

> use mean of half-hourly GWh otherwise the periods with more half-hours tend to visually dominate

```{r TheilSenGWhPlot, fig.cap=paste0("Theil-Sen plot of mean half-hourly generation per month since ", min(gridGenDT$rDateTimeUTC))}

gridGenDT[, date := as.Date(rDateTimeUTC)]  # set date to date for this one

# plotDT <- gridGenDT[, .(GWh = sum(GWh)), keyby = .(obsDate, peakPeriod)]
# plotDT[, date := obsDate]
# use mean of half-hourly GWh otherwise the periods with more half-hours tend to visually dominate
oaGWh <- openair::TheilSen(gridGenDT, "GWh", 
                  ylab = "GWh", deseason = TRUE,
                  xlab = "Year",
                  date.format = "%Y",
                  date.breaks = 4,
                  type = c("peakPeriod"), layout = c(5,1))
p <- oaGWh$plot
png(filename=here::here("docs/plots/openAir_UK_TheilSenGWhPlot.png"), 
    width = 960, units = "px")
# dpi = 600, width = 13.333, height = 7.5, units = "in"
plot(p)
dev.off()

t <- gridGenDT[, .("Mean GWh" = mean(GWh),
                   "N obs" = .N), keyby = .(year, peakPeriod)]

t <- dcast(t, year ~ peakPeriod, value.var = "Mean GWh")

kableExtra::kable(t, caption = "Mean half-hourly GWh by year and peak period - 2020 incomplete", digits = 2) %>%
  kable_styling()

```

Use the openair results object to estimate what the 2020 values would have been if the deaseasoned trend slope had continued.

> need to use the Theil-Sen slope model to estimate what the values would have been

```{r calculateGWhDiffFromTrend}
getModelTrendTable <- function(oa, fname){
  # oa is an openAir object created by theilSen
  # calculates the % below trend using the theil sen slope line parameters
  # oa <- oaGWh
  oaData <- as.data.table(oa$data$main.data)
  rDT <- oaData[, .(peakPeriod, date, conc, a,b,slope)]
  # https://github.com/davidcarslaw/openair/blob/master/R/TheilSen.R#L192
  # and
  # https://github.com/davidcarslaw/openair/blob/master/R/TheilSen.R#L625
  rDT[, x := time_length(date - as.Date("1970-01-01"), unit="days")] # n days since x = 0
  rDT[, expectedVal := a + (b * x)] # b = slope / 365
  
  # checks
  p <- ggplot2::ggplot(rDT, aes(x = date, colour = peakPeriod)) + 
    geom_line(aes(y = conc)) +
    labs(y = "Value",
         caption = fname)
  p <- p + geom_line(aes(y = expectedVal), linetype = "dashed")
  ggplot2::ggsave(here::here("docs", "plots", paste0("UK_trendModelTestPlot_", fname, ".png")))
  rDT[, diff := conc - expectedVal]
  rDT[, pcDiff := (diff/expectedVal)*100]
  
  t <- rDT[,.(peakPeriod, date, conc, a,b,slope,expectedVal, diff, pcDiff)]
  return(t)
}

t <- getModelTrendTable(oaGWh, "oaGWh")

ft <- dcast(t[date >= as.Date("2020-01-01")], date ~ peakPeriod, value.var = c("diff","pcDiff"))
ft[, date := format.Date(date, format = "%b %Y")]
ft[, diffTotal := `diff_Early morning` + `diff_Morning peak` + `diff_Day time` + 
     `diff_Evening peak` + `diff_Late evening`]
diffFT <- ft[, .(date, `diff_Early morning`, `diff_Morning peak`, `diff_Day time`, `diff_Evening peak`,
                 `diff_Late evening`, `diffTotal`)]
kableExtra::kable(diffFT, caption = "GWh above/below expected mean half-hourly GWh by month", digits = 2) %>%
  kable_styling()
pcFT <- ft[, .(date, `pcDiff_Early morning`, `pcDiff_Morning peak`, `pcDiff_Day time`, `pcDiff_Evening peak`,
                 `pcDiff_Late evening`)]
kableExtra::kable(pcFT, caption = "% above/below expected mean half-hourly GWh by month", digits = 2) %>%
  kable_styling()

```

Theil-Sen by weekday (for fun):

```{r theilSenGWhDoWPlot}
gridGenDT[, dow := lubridate::wday(obsDate, label = TRUE)]
openair::TheilSen(gridGenDT, "GWh", 
                  ylab = "GWh", deseason = TRUE,
                  xlab = "Year",
                  date.format = "%Y",
                  date.breaks = 4,
                  type = c("dow"), layout = c(7,1)) 
```

### Half hourly patterns

Figure \@ref(fig:recentDateTimeGWhPlot) shows total half-hourly generation since `r localParams$recentCutDate`. Overall generation has fallen as we would expect given the season (less heating and lighting required) and weekdays are indeed less easy to distinguish from weekends.

```{r recentDateTimeGWhPlot, fig.cap="Half-hourly generation, UK (recent)"}
# NB: the drake target, the rmd chunk and thus the crossref target are all the same!
p <- drake::readd(recentDateTimeGWhPlot)
yMax <- max(p$data$GWh) #RTFM
yMin <- min(p$data$GWh)
addVeDayDateTime(p, yMin = yMin, yMax = yMax)
```

Figure \@ref(fig:recentHalfHourlyProfileGWhPlot) shows the daily generation profiles over time for each day of the week. Clearly the shapes are both reducing in magnitude (seasonal and lockdown effects) and also converging in shape.

```{r recentHalfHourlyProfileGWhPlot, fig.cap="Half-hourly generation, UK (recent)"}
p <- drake::readd(recentHalfHourlyProfileGWhPlot)
p
```

```{r setCutYear}
cutYear <- 2018 # just compare with 2018 & 2019
compareYearsLabel <- "2018 & 2019"
alignedDT[, compareYear := ifelse(compareYear == 2020, compareYear,
                                  compareYearsLabel)]
gridGenDT[, month := lubridate::month(date)]
thisMonth <- lubridate::month(today())
```

Figure \@ref(fig:makeWeekDayGWhPlot) shows the mean half-hourly GWh per day. Need to bear in mind the trends identified above.

```{r makeWeekDayGWhPlot, fig.cap="Half-hourly GWh"}
dt <- alignedDT[dateFixed > as.Date("2020-01-01") & 
                                               dateFixed < localParams$toDate]
yVar <- "GWh"
xVar <- "wkdayFixed"
p <- makeWeekdayPlot(dt, xVar, yVar = yVar, yLab = yVar, yDiv = 1)

ggplot2::ggsave(filename = paste0("UK_weekdayPlot_",yVar,".png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p

```

Figure \@ref(fig:makeWeekdayTimeGWhPlot) shows the mean half-hourly CO2e per half-hour of each day. This plot shows the values as a proportion of the sum of the means across all half hours of each day to illustrate the relative, rather than absolute, shift in generation.

```{r makeWeekdayTimeGWhPlot, fig.cap="% half-hourly GWh"}

p <- makeWeekdayTimePlot(dt, yVar = yVar, yLab = "%", yDiv = 1000)

ggplot2::ggsave(filename = paste0("UK_weekdayTimePlot_",yVar,".png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```


Figure \@ref(fig:timeVariationGWhDetailed) shows the difference between hourly and weekday patterns for lockdown 2020 and the same period in all years from `r cutYear` onwards. It is interesting to note that the twin peak demand periods have been maintained during lockdown but they are considerably lower although we should expect some 'natural' reduction due to the overall downward generation trend shown in Figure \@ref(fig:TheilSenGWhPlot). The lower middle monthly plot suggests that while lockdown caused a considerable drop in generation, this difference is now starting to converge on 'normal' seasonal demand levels.

```{r timeVariationGWhDetailed, fig.cap=paste0("timeVariation plots for half-hourly GW generation comparing pre/lockdown 2020 with the same months in previous years starting in ", cutYear)}

t <- alignedDT[, .(min = min(rDateTimeUTC),
              max = max(rDateTimeUTC)), keyby = .(plotPeriodDetailed, compareYear)]

plotDT <- alignedDT[dateFixed > as.Date("2020-01-01") & dateFixed < today(),
                    .(rDateTimeUTC, GWh,compareYear,plotPeriodDetailed)]
plotDT[, date := rDateTimeUTC] # need to use original datetime as timeVariation expects unique dates

openair::timeVariation(plotDT, pollutant = "GWh", ylab = "GWh", # normalise = TRUE, not useful as divides by lockdown period mean so difference vanishes
                       group = "compareYear", type = "plotPeriodDetailed", 
                       key.columns = 2 # difference = TRUE,  fails: https://github.com/davidcarslaw/openAir/issues/166
                       )
```

Repeat but less detail.

```{r timeVariationGWh, fig.cap=paste0("timeVariation plots for half-hourly GWh generation comparing pre/lockdown 2020 up to ", localParams$toDate)}

t <- alignedDT[, .(min = min(rDateTimeUTC),
              max = max(rDateTimeUTC)), keyby = .(plotPeriod, compareYear)]


plotDT <- alignedDT[dateFixed > as.Date("2020-01-01") & 
                      dateFixed < localParams$toDate] # forces the comparison period
plotDT[, date := rDateTimeUTC]

out <- openair::timeVariation(plotDT, 
                       "GWh", ylab = "GWh", # normalise = TRUE, not useful as divides by lockdown period mean so difference vanishes
                       group = "compareYear", type = "plotPeriod")

message("Confirming dates plotted: date (as set for openair)")
summary(plotDT$date)
message("Confirming dates plotted: dateFixed")
summary(plotDT$dateFixed)
message("Confirming months plotted: date")
plotDT[, month := lubridate::month(date, lab = TRUE)]
table(plotDT$month)
```

```{r extractPlotsGWh}
# Extract the plots we want...
# do we get a legend?
p <- out$plot$day
png(filename=here::here("docs/plots/openAir_UK_DailyGWh.png"))
plot(p)
dev.off()
p <- out$plot$hour
png(filename=here::here("docs/plots/openAir_UK_HourlyGWh.png"))
plot(p)
dev.off()
p <- out$plot$day.hour
png(filename=here::here("docs/plots/openAir_UK_WeekdaysGWh.png"), width = 800, height = 600)
plot(p)
dev.off()
```

Figure \@ref(fig:timeVariationGWhNormalised) is a similar plot but in this case the generation is normalised within each period to enable clarification of changes in the temporal _shape_ of demand rather than changes in the overall level. As we can see the normalised plots indicate that the relative temporal distribution of generation (demand) is very similar between each period. The shape of demand is therefore relatively unchanged, but it's level is distinctly lower (Figure \@ref(fig:timeVariationGWh)).

```{r timeVariationGWhNormalised, fig.cap=paste0("timeVariation plots for normalised half-hourly GW generation comparing pre/lockdown 2020 with the same months in previous years starting in ", cutYear)}

openair::timeVariation(plotDT, 
                       "GW", ylab = "GW", normalise = TRUE, # 
                       group = "compareYear", type = "plotPeriod", 
                       key.columns = 2)
```


### Daily patterns

For this analysis we have shifted the dates for the comparison years to ensure that weekdays and weekends align in each year of data but this does **not** mean that Easter is the same weekend across the comparison periods nor that Bank Holidays (e.g. VE day 2020) are consistent.

Figure \@ref(fig:compareDailyGWhPlot) shows the most recent mean daily half-hourly GW compared to the same day in previous years. As hinted by Figure \@ref(fig:timeVariationGWh), lockdown clearly reduced demand below the seasonal trend as Figure \@ref(fig:TheilSenGWhPlot) suggested and also levels of demand are now starting to converge on the seasonal trend (c.f. Figure \@ref(fig:timeVariationGWh)).

> Beware temperature differences - need to control for heating degree days as 2020 was a _very_ warm spring

```{r compareDailyGWhPlot, fig.cap="Comparative daily mean half-hour generation levels 2020 vs pre-2020"}
p <- drake::readd(compareDailyGWhPlot)
yMax <- max(p$data$yMean) #RTFM
yMin <- min(p$data$yMean)
p <- addVeDayDate(p, yMin = yMin, yMax = yMax)
p <- addDSTBreakDate(p, yMin = yMin, yMax = yMax)
ggplot2::ggsave(filename = paste0("UK_dailyPlot_GWh.png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

Figure \@ref(fig:compareDailyGWhpcPlot) shows the percentage difference between the mean half-hourly generation per day in 2020 and the pre 2020 average for the same day. As we can see January 2020 was already slightly lower than previous years but February appears similar. There appears to be a substantial lockdown effect albiet with some fluctuations and very large dips on VE Day (Friday 8th May) and Thursday 28th May.

```{r compareDailyGWhpcPlot, fig.cap="Percentage difference in mean generation levels 2020 vs pre 2020"}

p <- drake::readd(compareDailyGWhpcPlot)
yMax <- max(p$data$pcDiffMean) #RTFM
yMin <- min(p$data$pcDiffMean)
p <- addVeDayDate(p, yMin = yMin, yMax = yMax)
p <- addDSTBreakDate(p, yMin = yMin, yMax = yMax)
ggplot2::ggsave(filename = paste0("UK_dailyPlot_GWh_pcCompare.png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

Figure \@ref(fig:peakyTest) shows the trend in the ratio between maximum and minimum half-hourly generation by day. This shows that 2020 was already less 'peaky' than previous years with a lower ratio between maximum and minimum generation on most days. This appear to have amplified in the first week or so of UK lockdown with a further reduction in the ratio but this has not been sustained over time with greater variance as we move into May 2020.

```{r peakyTest, fig.cap="Ratio of maximum to minimum generation per day"}
# calculate mean peakiness by relevant comparisons?
dailyDT <- alignedDT[dateFixed < lubridate::today(), .(maxGWh = max(GWh),
                     minGWh = min(GWh)
                     ), keyby = .(dateFixed, weekDay, wkdayFixed , compareYear)]
dailyDT[, peakyNess := maxGWh/minGWh] # 1 = no peak

yMax <- max(dailyDT$peakyNess)
yMin <- min(dailyDT$peakyNess)

p <- ggplot2::ggplot(dailyDT, aes(x = dateFixed, 
                                   y = peakyNess,
                                   shape = weekDay,
                                   colour = compareYear)) +
    geom_point() +
    geom_line(aes(shape = NULL), linetype = "dashed") + # joint the dots within compareYear
    scale_x_date(date_breaks = "7 day", date_labels =  "%a %d %b")  +
    theme(axis.text.x=element_text(angle=90, hjust=1)) +
    labs(caption = paste0(localParams$lockdownCap, localParams$weekendCap,
                          "\n", localParams$gamCap),
         x = "Date",
         y = "Ratio of max/min GWh"
    ) +
    theme(legend.position = "bottom") + 
    geom_smooth(aes(shape = NULL), method = "loess") + # will get a smooth line per year not per day
    scale_color_discrete(name = "Period") +
    scale_shape_discrete(name = "Weekday") +
    guides(colour=guide_legend(nrow=2)) +
    guides(shape=guide_legend(nrow=2))
  p <- addWeekendRectsDate(p, 
                           yMin, 
                           yMax
  )
p <- addLockdownRect(p, 
                       from = gcParams$UKlockDownStartDate, 
                       to = gcParams$UKlockDownEndDate, 
                       label = "Phase 1", yMin, yMax)
  
yMax <- max(p$data$peakyNess) #RTFM
yMin <- min(p$data$peakyNess)
addVeDayDate(p, yMin = yMin, yMax = yMax)
```


### Weekly patterns

tbc

## Carbon Intensity: Analysing deviation from 'normal'

There are two aspects to this. The first is `carbon intensity` which is driven by the mix of fuels being used to generate electricity at any given time. The second is the total greenhouse gasses emitted which is, obviously, the intensity * the volume. Given the slight uncertainty over units (see Section \@ref(data)) we assume this is `GENERATION` * `CARBON_INTENSITY`.

Clearly the first of these is driven by the mix of fuels and in the UK this reflects a complex dynamic system of availability of renewables, price, interconnect and demand. To some extent we would expect that lower overall demand should (but not [always](https://www.nationalgrideso.com/news/our-electricity-greener-lockdown)) increase the share of renewables. However other factors are also at play:

 * some electricity-usage practices may have shifted to (or indeed away from) periods which are likely to have high renewable availability;
 * some electricity-usage practices may have shifted away from the 'usual' morning and evening peak periods which are traditionally thought to require carbon intense peaking generation if sufficient pumped hydro is not available;
 * low periods of demand might [require system services](https://www.nationalgrideso.com/news/our-electricity-greener-lockdown) from higher carbon generation

### Overall trends

Figure \@ref(fig:TheilSenCIPlot) agin uses the openair `TheilSen()` function [@openair] to create a de-seasoned trend plot but this time for mean half-hourly carbon intensity by peak demand period. Again Table \@ref(tab:TheilSenCIPlot) indicates the years over which the trend is calculated. This plot shows that carbon intensity has fallen by around 40 kg CO2e/MWh per year over the time period in all time periods. However it is also noticeable that mean carbon intensity has marginally increased above trend in the months of lockdown.

```{r TheilSenCIPlot, fig.cap=paste0("Theil-Sen plot of mean half-hourly carbon intensity per month since ", min(gridGenDT$rDateTimeUTC))}
gridGenDT[, date := as.Date(rDateTimeUTC)]  # set date to date for this one

oaCI <- openair::TheilSen(gridGenDT, "CARBON_INTENSITY", 
                  ylab = "Carbon Intensity (kg CO2e/kWh)", deseason = TRUE,
                  xlab = "Year",
                  date.format = "%Y",
                  date.breaks = 4,
                  type = c("peakPeriod"), layout = c(5,1))

t <- gridGenDT[, .("Mean CI" = mean(CARBON_INTENSITY),
                   "N obs" = .N), keyby = .(year, peakPeriod)]

t <- dcast(t, year ~ peakPeriod, value.var = "Mean CI")
kableExtra::kable(t, digits = 2,
                  caption = "Mean half-hourly carbon intensity (kg CO2e/kWh) by year and peak period - 2020 incomplete") %>%
  kable_styling()

p <- oaCI$plot
png(filename=here::here("docs/plots/openAir_UK_TheilSenCIPlot.png"), 
    width = 960, units = "px")
# dpi = 600, width = 13.333, height = 7.5, units = "in"
plot(p)
dev.off()

```


Use the openair results object to estimate what the 2020 values would have been if the deaseasoned trend slope had continued. 

```{r calculateCIDiffFromTrend}

t <- getModelTrendTable(oaCI, "oaCI")

ft <- dcast(t[date >= as.Date("2020-01-01")], date ~ peakPeriod, value.var = c("diff","pcDiff"))
ft[, date := format.Date(date, format = "%b %Y")]
ft[, diffTotal := `diff_Early morning` + `diff_Morning peak` + `diff_Day time` + 
     `diff_Evening peak` + `diff_Late evening`]
diffFT <- ft[, .(date, `diff_Early morning`, `diff_Morning peak`, `diff_Day time`, `diff_Evening peak`,
                 `diff_Late evening`, `diffTotal`)]
kableExtra::kable(diffFT, caption = "GWh above/below expected mean half-hourly CI by month", digits = 2) %>%
  kable_styling()
pcFT <- ft[, .(date, `pcDiff_Early morning`, `pcDiff_Morning peak`, `pcDiff_Day time`, `pcDiff_Evening peak`,
                 `pcDiff_Late evening`)]
kableExtra::kable(pcFT, caption = "% above/below expected mean half-hourly CI by month", digits = 2) %>%
  kable_styling()

```

Theil-Sen by weekday (for fun):

```{r theilSenCIDoWPlot}

openair::TheilSen(gridGenDT, "CARBON_INTENSITY", 
                  ylab = "g CO2e/kWh", deseason = TRUE,
                  xlab = "Year",
                  date.format = "%Y",
                  date.breaks = 4,
                  type = c("dow"), layout = c(7,1)) 
```

### Half-hourly patterns

Figure \@ref(fig:recentDateTimeCIPlot) shows half-hourly carbon intensity since `r localParams$recentCutDate`. Overall generation has fallen as we would expect given the season (less heating and lighting required) and weekdays are indeed less easy to distinguish from weekends.

```{r recentDateTimeCIPlot, fig.cap="Half-hourly CO2e emissions, UK (recent)"}
p <- drake::readd(recentDateTimeCIPlot)
yMax <- max(p$data$yVals) #RTFM
yMin <- min(p$data$yVals)
addVeDayDateTime(p, yMin = yMin, yMax = yMax)
```

Figure \@ref(fig:recentHalfHourlyProfileCIPlot) shows the daily carbon intensity profiles over time for each day of the week.

```{r recentHalfHourlyProfileCIPlot, fig.cap="Half-hourly generation, UK (recent)"}
p <- drake::readd(recentHalfHourlyProfileCIPlot)
p
```

Figure \@ref(fig:makeWeekDayCIPlot) shows the mean half-hourly CO2e per day. Need to bear in mind the trends identified above.

```{r makeWeekDayCIPlot, fig.cap="half-hourly CI"}
dt <- alignedDT[dateFixed > as.Date("2020-01-01") & 
                                               dateFixed < localParams$toDate]
yVar <- "CARBON_INTENSITY"
xVar <- "wkdayFixed"
p <- makeWeekdayPlot(dt, xVar, yVar = yVar, yLab = "g C02e/kWh", yDiv = 1)

ggplot2::ggsave(filename = paste0("UK_weekdayPlot_",yVar,".png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

Figure \@ref(fig:makeWeekdayTimeCIplot) shows the mean half-hourly CO2e per half-hour of each day. This plot shows the values as a proportion of the sum of the means across all half hours of each day to illustrate the relative, rather than absolute, shift in generation.

```{r makeWeekdayTimeCIplot, fig.cap="% half-hourly CI"}

p <- makeWeekdayTimePlot(dt, yVar = yVar, yLab = "%", yDiv = 1000)

ggplot2::ggsave(filename = paste0("UK_weekdayTimePlot_",yVar,".png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```


Figure \@ref(fig:timeVariationCI) shows the difference between hourly and weekday patterns for lockdown 2020 and the previous year(s) starting from January `r cutYear`. This plot is not particularly informative since we now CI is already lower in 2020 than previous years and we would expect it to fall during the spring as solar generation increases its contributon. The lockdown phases will also be affected by small numbers of highly windy days.

```{r timeVariationCI, fig.cap=paste0("timeVariation plots for half-hourly carbon intensity comparing lockdown 2020 up to ", localParams$toDate)}


plotDT <- alignedDT[dateFixed > as.Date("2020-01-01") & 
                                               dateFixed < localParams$toDate]

plotDT[, date := rDateTimeUTC]

out <- openair::timeVariation(plotDT, 
                       "CARBON_INTENSITY", ylab = "CO2e kg/kWh", # normalise = TRUE, not useful as divides by lockdown period mean so difference vanishes
                       group = "compareYear", type = "plotPeriod", 
                       key.columns = 2)

message("Confirming dates plotted: date (as set for openair)")
summary(plotDT$date)
message("Confirming dates plotted: dateFixed")
summary(plotDT$dateFixed)
message("Confirming months plotted: date")
plotDT[, month := lubridate::month(date, lab = TRUE)]
table(plotDT$month)
```

```{r extractPlotsCI}
# Extract the plots we want...
p <- out$plot$day
png(filename=here::here("docs/plots/openAir_UK_DailyCI.png"))
plot(p)
dev.off()
p <- out$plot$hour
png(filename=here::here("docs/plots/openAir_UK_HourlyCI.png"))
plot(p)
dev.off()
p <- out$plot$day.hour
png(filename=here::here("docs/plots/openAir_UK_WeekdaysCI.png"), width = 800, height = 600)
plot(p)
dev.off()
```

### Daily patterns

Figure \@ref(fig:compareDailyCIPlot) shows the mean half-hourly carbon intensity per day in 2020 and the pre 2020 average for the same day. As we would expect given Figure \@ref(fig:TheilSenCIPlot), 2020 was already considerably lower than the average of previous years but this is not necessarily sustained through lockdown although the affects of weather on solar and wind availability need to be taken in to account.

```{r compareDailyCIPlot, fig.cap="Percentage difference in mean carbon intensity levels 2020 vs pre 2020"}
p <- drake::readd(compareDailyCIPlot)
yMax <- max(p$data$yMean) #RTFM
yMin <- min(p$data$yMean)
p <- addVeDayDate(p, yMin = yMin, yMax = yMax)
p <- addDSTBreakDate(p, yMin = yMin, yMax = yMax)
ggplot2::ggsave(filename = paste0("UK_dailyPlot_CI.png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

Figure \@ref(fig:compareDailyCIpcPlot) shows the percentage difference between the mean half-hourly carbon intensity per day in 2020 and the pre 2020 average for the same day. As expected, 2020 was already considerably lower than the average of previous years but this is not necessarily sustained through lockdown although the affects of weather on solar and wind availability need to be taken in to account.

```{r compareDailyCIpcPlot, fig.cap="Percentage difference in mean carbon intensity levels 2020 vs pre 2020"}
p <- drake::readd(compareDailyCIpcPlot)
yMax <- max(p$data$pcDiffMean) #RTFM
yMin <- min(p$data$pcDiffMean)
p <- addVeDayDate(p, yMin = yMin, yMax = yMax)
p <- addDSTBreakDate(p, yMin = yMin, yMax = yMax)
ggplot2::ggsave(filename = paste0("UK_dailyPlot_CI_pcCompare.png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

### Weekly patterns

tbc

## Carbon emissions (total): Analysing deviation from 'normal'

In this section we use our calculation of total CO2e emitted per half hour (see Section \@ref(genData)) to analyse the changes in total CO2e emitted which is, after all, what we are mostly interested in from a climate change point of view. Remember that this value is driven _both_ by total generation (demand) _and_ carbon intensity. As we saw above, these are not always tightly correlated.

### Overall trends

Figure \@ref(fig:TheilSenC02eTPlot) again uses the openair `TheilSen()` function [@openair] to create a de-seasoned trend plot but this time for mean half-hourly carbon intensity by peak demnand period. Again Table \@ref(tab:TheilSenC02eTPlot) indicates the years over which the trend is calculated. 

This plot shows that total CO2e emissions intensity has fallen by a half-hourly mean of ~600-800 tonnes of CO2e over the time period. However there is some variation around this trend with apparently slightly above trend emissions in the evening peak period during April and May 2020 noticeable despite the almost complete absence of coal generation during this period (see Annex, Figure \@ref(fig:fuelTrendPlots)).

```{r TheilSenC02eTPlot, fig.cap=paste0("Theil-Sen plot of mean half-hourly kg CO2e emissions per month since ", min(gridGenDT$rDateTimeUTC))}
gridGenDT[, date := as.Date(rDateTimeUTC)]  # set date to date for this one

oaCO2 <- openair::TheilSen(gridGenDT, "C02e_T", 
                  ylab = "Mean half-hourly C02e (T)", deseason = TRUE,
                  xlab = "Year",
                  date.format = "%Y",
                  date.breaks = 4,
                  type = c("peakPeriod"), layout = c(5,1))
p <- oaCO2$plot
png(filename=here::here("docs/plots/openAir_UK_TheilSenTotalCO2ePlot.png"), 
    width = 960, units = "px")
# dpi = 600, width = 13.333, height = 7.5, units = "in"
plot(p)
dev.off()

t <- gridGenDT[, .("Mean half-hourly C02e (T)" = mean(C02e_T),
                   "N obs" = .N), keyby = .(year, peakPeriod)]

t <- dcast(t, year ~ peakPeriod, value.var = "Mean half-hourly C02e (T)")

kableExtra::kable(t, caption = "Mean half-hourly C02e (T) by year and peak period - 2020 incomplete", 
                  digits = 2) %>%
  kable_styling()
```

Now use the openair results object to estimate what the 2020 values would have been if the deaseasoned trend slope had continued. 

```{r calculateCO2DiffFromTrend}

t <- getModelTrendTable(oaCO2, "oaCO2")

ft <- dcast(t[date >= as.Date("2020-01-01")], date ~ peakPeriod, value.var = c("diff","pcDiff"))
ft[, date := format.Date(date, format = "%b %Y")]
ft[, diffTotal := `diff_Early morning` + `diff_Morning peak` + `diff_Day time` + 
     `diff_Evening peak` + `diff_Late evening`]
diffFT <- ft[, .(date, `diff_Early morning`, `diff_Morning peak`, `diff_Day time`, `diff_Evening peak`,
                 `diff_Late evening`, `diffTotal`)]
kableExtra::kable(diffFT, caption = "GWh above/below expected mean half-hourly CO2e (T) by month", digits = 2) %>%
  kable_styling()
pcFT <- ft[, .(date, `pcDiff_Early morning`, `pcDiff_Morning peak`, `pcDiff_Day time`, `pcDiff_Evening peak`,
                 `pcDiff_Late evening`)]
kableExtra::kable(pcFT, caption = "% above/below expected mean half-hourly CO2e (T) by month", digits = 2) %>%
  kable_styling()
```

Theil-Sen by weekday (for fun):

```{r theilSenCO2eDoWPlot}
openair::TheilSen(gridGenDT, "C02e_T", 
                  ylab = "CO2e (T)", deseason = TRUE,
                  xlab = "Year",
                  date.format = "%Y",
                  date.breaks = 4,
                  type = c("dow"), layout = c(7,1)) 
```

### Half-hourly patterns

Figure \@ref(fig:recentDateTimeC02ekgPlot) shows total half-hourly CO2e emissions since `r localParams$recentCutDate`. Overall generation has fallen as we would expect given the season (less heating and lighting required) and weekdays are indeed less easy to distinguish from weekends.

```{r recentDateTimeC02ekgPlot, fig.cap="Half-hourly total CO2e emissions, UK (recent)"}
p <- drake::readd(recentDateTimeC02ekgPlot)
yMax <- max(p$data$yVals) #RTFM
yMin <- min(p$data$yVals)
addVeDayDateTime(p, yMin = yMin, yMax = yMax)
```

Figure \@ref(fig:recentHalfHourlyProfileC02ekgPlot) shows the daily CO2e emissions profiles over time for each day of the week since `r localParams$recentCutDate`.

```{r recentHalfHourlyProfileC02ekgPlot, fig.cap="Half-hourly CO2e emissions, UK (recent)"}
p <- drake::readd(recentHalfHourlyProfileC02ekgPlot)
p
```

Figure \@ref(fig:makeWeekDayCO2eTPlot) shows the mean half-hourly CO2e per day. Need to bear in mind the trends identified above.

```{r makeWeekDayCO2eTPlot, fig.cap="half-hourly CO2e"}
dt <- alignedDT[dateFixed > as.Date("2020-01-01") & 
                                               dateFixed < localParams$toDate]
yVar <- "C02e_T"
xVar <- "wkdayFixed"
p <- makeWeekdayPlot(dt, xVar, yVar = yVar, yLab = "C02e (kT)", yDiv = 1000)

ggplot2::ggsave(filename = paste0("UK_weekdayPlot_",yVar,".png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

Figure \@ref(fig:makeWeekdayTimeCO2eTplot) shows the mean half-hourly CO2e per half-hour of each day. This plot shows the values as a proportion of the sum of the means across all half hours of each day to illustrate the relative, rather than absolute, shift in generation.

```{r makeWeekdayTimeCO2eTplot, fig.cap="% half-hourly CO2e"}

p <- makeWeekdayTimePlot(dt, yVar = yVar, yLab = "%", yDiv = 1000)

ggplot2::ggsave(filename = paste0("UK_weekdayTimePlot_",yVar,".png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

Figure \@ref(fig:timeVariationCO2e) shows the difference between hourly and weekday patterns for lockdown 2020 and the previous year(s) starting from January `r cutYear`. Again, due to the downward trends we have already identified, we would expect total CO2e emissions to be lower during 2020 and also to fall during the spring as solar generation increases its contributon.

```{r timeVariationCO2e, fig.cap=paste0("timeVariation plots for half-hourly CO2e comparing lockdown 2020 with pre-lockdown starting in ", cutYear)}

plotDT <- alignedDT[dateFixed > as.Date("2020-01-01") & 
                                               dateFixed < localParams$toDate]

plotDT[, date := rDateTimeUTC]

out <- openair::timeVariation(plotDT, 
                       "C02e_T", ylab = "Mean half-hourly C02e (T)", # normalise = TRUE, not useful as divides by lockdown period mean so difference vanishes
                       group = "compareYear", type = "plotPeriod", 
                       key.columns = 2)

message("Confirming dates plotted: date (as set for openair)")
summary(plotDT$date)
message("Confirming dates plotted: dateFixed")
summary(plotDT$dateFixed)
message("Confirming months plotted: date")
plotDT[, month := lubridate::month(date, lab = TRUE)]
table(plotDT$month)
```

```{r extractPlotsCO2e_T}
# Extract the plots we want...
p <- out$plot$day
png(filename=here::here("docs/plots/openAir_UK_DailyCO2e_T.png"))
plot(p)
dev.off()
p <- out$plot$hour
png(filename=here::here("docs/plots/openAir_UK_HourlyCO2e_T.png"))
plot(p)
dev.off()
p <- out$plot$day.hour
png(filename=here::here("docs/plots/openAir_UK_WeekdaysCO2e_T.png"), width = 800, height = 600)
plot(p)
dev.off()
```

### Daily patterns

Figure \@ref(fig:compareDailyCO2ekgPlot) shows the mean half-hourly carbon intensity per day in 2020 and the pre 2020 average for the same day. As expected, 2020 was already considerably lower than the average of previous years but this is not necessarily sustained through lockdown although the affects of weather on solar and wind availability need to be taken in to account.

> check y axis

```{r compareDailyCO2ekgPlot, fig.cap="Mean half-hourly total CO2e by day, 2020 vs pre 2020"}
p <- drake::readd(compareDailyCO2ekgPlot)
yMax <- max(p$data$yMean) #RTFM
yMin <- min(p$data$yMean)
p <- addVeDayDate(p, yMin = yMin, yMax = yMax)
p <- addDSTBreakDate(p, yMin = yMin, yMax = yMax)
ggplot2::ggsave(filename = paste0("UK_dailyPlot_CO2e.png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

Figure \@ref(fig:compareDailyC02ekgpcPlot) shows the percentage difference between the mean half-hourly carbon intensity per day in 2020 and the pre 2020 average for the same day. As expected, 2020 was already considerably lower than the average of previous years but this is not necessarily sustained through lockdown although the affects of weather on solar and wind availability need to be taken in to account.

```{r compareDailyC02ekgpcPlot, fig.cap="Percentage difference in mean half-hourly CO2e 2020 vs pre 2020"}
p <- drake::readd(compareDailyC02ekgpcPlot)
yMax <- max(p$data$pcDiffMean) #RTFM
yMin <- min(p$data$pcDiffMean)
p <- addVeDayDate(p, yMin = yMin, yMax = yMax)
p <- addDSTBreakDate(p, yMin = yMin, yMax = yMax)
ggplot2::ggsave(filename = paste0("UK_dailyPlot_CO2e_pcCompare.png"), 
                  path = here::here("docs", "plots"), 
                  plot = p, width = 9, units = "in")
p
```

### Weekly patterns

tbc

# Summary (to date)

# About

## Citation

```{r citation, child=gcParams$citationUoS}
```

## Report circulation

 * Public – this report is intended for publication.

## Code

All code used to create this report is available from:

 * https://github.com/CfSOtago/gridCarbon

## License {#license}

```{r ccby license, child=gcParams$licenseCCBY}
```

## History

```{r history, child=gcParams$history}
```
 * this [report's edit history](https://github.com/CfSOtago/gridCarbon/commits/master/dataAnalysis/)
 
## Support

```{r generic support, child=gcParams$support}
```
 * The European Union via [SPATIALEC](http://www.energy.soton.ac.uk/tag/spatialec/), a Marie Skłodowska-Curie [Global Fellowship](https://ec.europa.eu/research/mariecurieactions/actions/individual-fellowships_en) based at the University of Otago’s [Centre for Sustainability](http://www.otago.ac.nz/centre-sustainability/staff/otago673896.html) (2017-2019) & the University of Southampton’s [Sustainable Energy Research Group](http://www.otago.ac.nz/centre-sustainability/staff/otago673896.html) (2019-2020) (Anderson)
 
\newpage

# Annexes

## Grid generation data

```{r skimGrid}
skimr::skim(gridGenDT)
```

## Fuel trends

```{r fuelGWByYear}
t <- gridGenDT[, .("coal" = mean(COAL),
                   "gas" = mean(GAS),
                   "nuclear" = mean(NUCLEAR),
                   "wind" = mean(WIND),
                   "hydro" = mean(HYDRO),
                   "solar" = mean(SOLAR),
                   "biomass" = mean(BIOMASS),
                   "imports" = mean(IMPORTS),
                   "other" = mean(OTHER),
                   "N obs" = .N), keyby = .(year)]

kableExtra::kable(t, caption = "Mean half-hourly GW by fuel by year", digits = 2) %>%
  kable_styling()
```

Figure \@ref(fig:fuelTrendPlots) shows mean half-hourly generation per month by fuel together with a smoothed trend line. Note that these values are _not_ de-seasoned.

```{r fuelTrendPlots, fig.cap=paste0("Mean half-hourly generation (GW) by fuel use per month since ", min(gridGenDT$rDateTimeUTC))}
gridGenDT[, date := as.Date(rDateTimeUTC)]  # set date to date for this one

makeTSFuelTrends <- function(fuels){
  for(f in fuels){
   openair::TheilSen(gridGenDT, eval(f), 
                  ylab = paste0(eval(f), " (GWh)") # deseason = TRUE can make things go < 0
                  ) 
  }
}

fuels <-c("COAL","GAS","NUCLEAR","WIND","HYDRO","BIOMASS","SOLAR","STORAGE","IMPORTS","OTHER")

# create the fuel trend plots
#makeTSFuelTrends(fuels)

```

Clearly some of these trends are non-monotonic and so are poorly represented by the Theil-Sen function. The next set of plots repeat this analysis using openair's smoothTrend plot which implemets a GAM-based method similar to \@ref(fig:compareDailyGWhPlot).

```{r smoothTrendPlots, fig.cap=paste0("Trend plots of mean half-hourly fuel use per month since ", min(gridGenDT$rDateTimeUTC))}
gridGenDT[, date := as.Date(rDateTimeUTC)]  # set date to date for this one

makeSmoothFuelTrends <- function(fuels){
  for(f in fuels){
   openair::smoothTrend(gridGenDT, eval(f), 
                  ylab = paste0(eval(f), " (MW)") # deseason = TRUE can make thjings go < 0
                  ) 
  }
}

fuels <-c("COAL","GAS","NUCLEAR","WIND","HYDRO","BIOMASS","SOLAR","STORAGE","IMPORTS","OTHER")

# create the fuel trend plots
makeSmoothFuelTrends(fuels)

```

## Experimental plots

Figure \@ref(fig:binedDemand) shows the relationship between half-hourly carbon intensity and half-hourly generation over the last five years. 2020 is excluded as it is not yet complete.

```{r binedDemand, fig.cap="Box & whiskers plots of carbon intensity over time by generation decile"}
library(gtools)

gridGenDT[, genBin := gtools::quantcut(GENERATION/1000, q=10)]

ggplot2::ggplot(gridGenDT[year < 2020], aes(x = genBin, colour = genBin,
                                            y = CARBON_INTENSITY)) +
  geom_boxplot() +
  theme(legend.position="bottom") +
  labs(x = "Generation decile (GW)",
       y = "CI KgCO2e/kWh") + 
  theme(axis.text.x=element_blank()) +
  scale_colour_viridis_d(name="GW") +
  facet_grid(. ~ year)
```

The openair package [@openair] contains a number of novel plotting methods.

```{r carbonRose}
# polluation rose but with pollutant = GW and wind direction = CI (as a % of 360)

oaDT <- gridGenDT[, .(rDateTimeUTC, ws = GENERATION/1000, CI = CARBON_INTENSITY)]
oaDT <- oaDT[, wd := 360 * (CI / max(CI))]

openair::windRose(oaDT)
```


```{r carbonScatter}
openair::scatterPlot(gridGenDT[year < 2020], x = "GENERATION", y = "CARBON_INTENSITY", 
                     method = "hexbin", col = "jet", type = c("year","season"))
```


# Runtime

```{r check runtime, include=FALSE}
t <- proc.time() - startTime
elapsed <- t[[3]]
```

Analysis completed in `r round(elapsed,2)` seconds ( `r round(elapsed/60,2)` minutes) using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform`.

# R environment

## R packages used

 * base R [@baseR]
 * bookdown [@bookdown]
 * data.table [@data.table]
 * drake [@drake]
 * ggplot2 [@ggplot2]
 * kableExtra [@kableExtra]
 * knitr [@knitr]
 * lubridate [@lubridate]
 * rmarkdown [@rmarkdown]
 * zoo [@zoo]

## Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References
